A flexible framework to prototype, develop and manage ubiquitous applications for intelligent environments

### 2013-09-20 trunk/ labtempo-ubicomp-projects/UbicompMiddleware/doc/PerCom2014/percom2014_conf.tex
Added a new paragraph at IPGAP section

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% \title{Interacting with the environment: a framework to develop ubiquitous applications for intelligent environments}
\title{Encapsulating the environment: a flexible framework to develop ubiquitous applications for intelligent environments}

@@

\begin{abstract}
%\boldmath
Due to recent advances in mobile computing and wireless communication technologies, we can see the emergence of a favorable scenario for building ubiquitous/pervasive applications. This work proposes a new framework that aims at building those applications, providing a set of concepts and implementation tools. We propose abstractions that allow developers to handle the resources spread in the environment in a simple and homogeneous way, and to interpret context information. In order to demonstrate the feasibility of the proposal, the concepts were implemented in a platform named \textit{SmartAndroid}. A ubiquitous applications prototyping interface was implemented over this platform to allow testing of different environment configurations before purchasing all devices; and also a context rules composition interface, where end users can define their preferences in the environment.

@@

%FIXME: araujo é um artigo em PT, arrumar um em EN
In this work we propose a framework for developing ubiquitous applications in IE. The goal is to provide support for the programming, testing and execution of applications, thus allowing to deal consistently with systems great complexity. The framework stands out for addressing the challenges already identified in UbiComp~\cite{araujo2003}. The \textit{heterogeneity of devices} is handled through the definition of a Distributed Component Model, that provides abstractions to encapsulate these devices, also called resources, enabling developers to interact with them seamlessly. Regarding the \textit{variety of context information and services} issue, we propose a solution for context interpretation that allows developers to create and manage context rules at runtime, and users to set their preferences in the IE. Finally, regarding the issue on the \textit{resource availability}, the framework includes an application interface for the prototyping and management of pervasive applications (IPGAP), focused on visualization and testing ubiquitous applications, mixing real and virtual components.
%in which the basic component has a uniform structure defined as a Resource Agent (RA), formerly proposed in~\cite{cardoso2006} as an entity gathering context information. In this work, the RA, while maintaining context information, acts as a component that encapsulates the device code associated with it, including aspects of interaction with application components. 
%The \textit{heterogeneity of devices} is handled through the definition of a Distributed Component Model, in which the basic component has a uniform structure defined as a Resource Agent (RA), formerly proposed in~\cite{cardoso2006} as an entity gathering context information. In this work, the RA, while maintaining context information, acts as a component that encapsulates the device code associated with it, including aspects of interaction with application components.

The concepts of the framework were materialized on a platform called \textit{SmartAndroid}\footnote{www.tempo.uff.br/smartandroid}, whose development has enabled appraisal in order to prove that the conceptual framework facilitates the process of building ubiquitous applications. We implemented some use cases to validate different aspects of the framework proposed and to testify its capabilities, supported by the elaboration of competency questions, which is a mechanism mostly used to evaluate ontologies, but can also be used as an assessment to this work. 
%power of representation 
%testify the feasibility 
% has enabled the construction of use cases that demonstrate the effectiveness of the approach to provide context awareness to new applications.

% The concepts of the framework have been implemented on a platform called \textit{SmartAndroid}\footnote{www.tempo.uff.br/smartandroid} developed in the context of the project. The implementation of this project has enabled appraisal in order to prove that the conceptual framework facilitates the process of building ubiquitous applications.  One of the evaluations occurred during the processing of an application running on a purely local ubiquitous application. In another assessment, an application was built to verify the feasibility of implementation of Context Model, through the exploitation of communication mechanisms used in the Distributed Component Model.

The remainder of this paper is organized as follows. First we present in Section~\ref{sec:overview} an overview of the main concepts used as a basis for the framework's development. In Section~\ref{sec:framework}, we present the framework's overall architecture and main features. We then, in Section~\ref{sec:evaluation}, present a proof of concept demonstrating the feasibility of building ubiquitous applications using the framework and we show examples of applications that explores the key features of the IE. Finally, in Section~\ref{sec:related} we compare our approach with related work and in Section~\ref{sec:conclusion} we conclude this paper with the main remarks.

\section{Related Work} \label{sec:related}
% [ubicomp is nice but it is not priceless]
The benefits to users and developers that UbiComp foresees reach beyond the horizon of many imaginative researchers. The possibilities that arise from it surpass the bounds established by keyboards and mice, our standard interfaces. Nonetheless, building such adaptable applications requires much effort from developers that see themselves immersed in a universe of device's specifications and communication technologies, apart from the problem itself that the application must solve.

%ubicomp implies dealing with limited and dinamically varying computational resources [kindberg2002ieee]

Therefore, many researchers have focused on diminishing those obstacles by providing abstractions, middlewares, services, tools, and other suportive techniques, not only for development but also for prototyping.

As one of the most famous, the Gaia middleware is designed to facilitate the construction of applications for IE. It consists of a set of core services and a framework for building distributed context aware applications. 
This project 
%uses operating systems abstractions to define its architecture
%have many goals that include the provision of mechanisms of acquisition of context; composition of rules (using first logic order), 
%%maintenance of hardwares and softwares description and about applications; 
%search mechanism to find resources
%presence and execution status of resources
%interpretation mechanism that gathers and interprets informations, as defined by programmers
%mvc
%The main goals of Gaia's approach are the acquisition of context by applications, the monitoring of entities location, the maintenance of hardwares and softwares descriptions, and 
%FIXME: second phrase was copied
%Active Spaces
%ontologies

@@

% [explain that our solution strives to accomodate several problems of UbiComp and to be flexible enough to accomodate the others]

\section{Overview} \label{sec:overview}
% [context/contx sensibility]
It is easy to notice that the context information is a first concern topic in ubiquitous systems. But what is ``context''? Many authors have proposed definitions to this concept, though most are little accurate~\cite{baldauf2007survey}. Synthetizing previous definitions, Dey and Abowd have proposed that context is ``any information that can be used to characterize the situation of entities (i.e., whether a person, place or object) that are considered relevant to the interaction between a user and an application, including the user and the application themselves''~\cite{Dey2001}. In a IE ``Places'' are the rooms, the floors of a building, or spaces in general that enable the localization of other entities, ``People'' are individuals that populate the environment and interact with it, and ``things'' are virtual representations of physical objects or software components.
%Considering the diversity of devices, systems and protocols, one 
%Many authors have proposed definitions to this concept, though most use examples to explain it or adopt a restrictive definition, that may not contemplate a set of systems. <- wrong. Most are wide or use synonyms
%Synthetizing previous definitions, Dey and Abowd have proposed that context is any relevant information used to caracterize the situation of entities, specifically: people, places and things.

Context aware (or context sensible) applications is a designation to those applications that not only are capable of knowing the context of the environment, but also react to it either by means of changing the environment or in a software level.

[distribs sys]

\section{Framework} \label{sec:framework}
[framework description]

\subsection{Resource Agents}

[communication mechanisms]

\subsection{Architeture}
[distributed components model]

\subsection{Management Services}
[resource register service]

[resource discovery service]

[resource location service]

[location concerns]

[security concerns]

\subsection{Context Interpretation}
[directly programmed]

[as a separated entity]

\subsection{Framework Implementation}
[android and other techs]

[impl of AR and services]

[impl of context rules]

[context rules conflicts is an ongoing study that aims at discover a conflict in runtime, ]

\subsection{Management and Prototyping Interface}
% interface de prototipagem e gerenciamento de aplicações pervasivas
% prototyping and management of pervasive applications interface
% pervasive applications' management and prototyping interface
% PAMPI
% management and prototyping interface MPI
[description]

>>>>

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor UbiApp UbiComp}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% \title{Interacting with the environment: a framework to develop ubiquitous applications for intelligent environments}
% \title{Encapsulating the environment: a flexible framework to develop ubiquitous applications for intelligent environments}
% \title{A flexible framework to develop ubiquitous applications for intelligent environments}
\title{A flexible framework to prototype, develop and manage ubiquitous applications for intelligent environments}

@@

\begin{abstract}
%\boldmath
% version 2
% Due to recent advances in mobile computing and wireless communication technologies, we can see the emergence of a favorable scenario for building ubiquitous/pervasive applications. This work proposes a framework that aims at building those applications, providing a set of concepts and implementation tools. We propose abstractions that allow developers to handle the resources spread in the environment in a simple and homogeneous way, and to interpret context information. In order to demonstrate the feasibility of the proposal, the concepts were implemented in a platform named \textit{SmartAndroid}. A ubiquitous applications prototyping interface was implemented over this platform to allow testing of different environment configurations before purchasing all devices; and also a context rules composition interface, where end users can define their preferences in the environment.

% version 3
% Due to recent advances in mobile computing and wireless communication technologies, we can see the emergence of a favorable scenario for building ubiquitous/pervasive applications. This work proposes a state-of-art framework for building those applications by solving some of the main issues of the field, namely the heterogeneity of devices, the amount and diversity of context informations and services, and the difficulty of developing and testing ubiquitous applications. The abstractions provided, and the management services conceived, allow developers to handle resources spread within the intelligent environment in a simple and homogeneous way. Furthermore, they enhance the capability of implementing context-aware applications, while still keep the flexibility on developing such applications. In order to demonstrate the applicability of the proposal, we implemented the concepts in a platform named \textit{SmartAndroid}, which contemplates most of the framework's features and has enabled the construction of a set of example applications to validate the framework. Above the platform we implemented a management and prototyping interface that uses a tablet device to allow the testing of applications running on different configurations of the environment. This intelligent environment may be populated with physical and/or virtual resources, what enables the emulation of devices as well as the simulation, without the need of purchasing all those devices. The interface also allows the composition of context rules either by developers or non trained users.

% version 4
Due to recent advances in mobile computing and wireless communication technologies, we can see the emergence of a favorable scenario for building ubiquitous/pervasive applications. This work proposes a state-of-art framework for building those applications by solving some of the main issues of the field, namely the heterogeneity of devices, the amount and diversity of context informations and services, and the difficulty of developing and testing ubiquitous applications. The abstractions provided, and the management services conceived, allow developers to handle resources spread within the intelligent environment in a simple and homogeneous way. Furthermore, they enhance the capability of implementing context-aware applications, while still keep the flexibility on developing such applications. In order to demonstrate the applicability of the proposal, we implemented the concepts in a platform named \textit{SmartAndroid}, which contemplates most of the framework's features and has enabled the construction of a set of example applications to validate the framework. Above the platform we implemented a management and prototyping interface that uses a tablet device to allow the testing and evaluation of applications running on different configurations of the environment. This intelligent environment may be populated with a mix of simulated and real devices, what may reduce the costs of prototyping. The interface also enables the end-user programming of context rules through a high level user interface.
% Above the platform we have developed a management and prototyping interface that uses a mobile device such as a tablet to allow the testing and evaluation of ubiquitous applications. Therefore it is possible to execute a wide range of scenarios, using a mix of simulated and real devices. This interface also permits end-user programming of context rules through a hi-level user interface.

% This intelligent environment may be populated with physical and/or virtual resources, what enables the emulation of devices as well as the simulation, without the need of purchasing all those devices. 
% 
% The interface also allows the composition of context rules either by developers or non trained users.
% 
% In addition, we have developed a management and prototyping interface, which runs in a mobile device such as a tablet, that allows testing and evaluation of ubiquitous applications. 
% 
% Therefore it is possible to execute a wide variety of scenarios, using a mix of simulated and real devices. This interface also permits end-user programming of context rules through a hi-level user interface.

% This intelligent environment may be loaded with the representation of physical resources and/or simulators, what enables the prototyping of applications without the need of purchasing all those devices.

% This intelligent environment may be loaded with the representation of physical and/or virtual resources, what enables the prototyping of applications without the need of purchasing all those devices.

% With the purpose of evaluating the proposal we implemented the concepts 
% and enable the context interpretation and 
%  even without the need of purchasing all those devices
%  (with a touch screen interface)
% The abstractions provided, and the management services conceived, allow developers to handle resources spread within the intelligent environment in a simple and homogeneous way, and enhances the capability of implementing context sensible applications, while still keep the flexibility on developing such applications. 
% Above the platform we implemented a management and prototyping interface that uses a tablet device to allow the testing of applications running on different configurations of the environment, which may be loaded with simulators or the representation of the resources. The interface also allows the composition of context rules either by developers or non trained users.
%FIXME: ``while still keep the flexibility ON developing such applications'' <- este on está certo?

% This work proposes a framework that aims at building those applications, providing a set of concepts and implementation tools. We propose abstractions that allow developers to handle the resources spread in the environment in a simple and homogeneous way, and to interpret context information. In order to demonstrate the feasibility of the proposal, the concepts were implemented in a platform named \textit{SmartAndroid}. A ubiquitous applications prototyping interface was implemented over this platform to allow testing of different environment configurations before purchasing all devices; and also a context rules composition interface, where end users can define their preferences in the environment.

@@

%FIXME: araujo é um artigo em PT, arrumar um em EN
In this work we propose a framework for developing UbiApps in IE. The goal is to provide support for the programming, testing and execution of applications, thus allowing to deal consistently with systems great complexity. The framework stands out for addressing the challenges already identified in UbiComp~\cite{araujo2003}. The \textit{heterogeneity of devices} is handled through the definition of a Distributed Component Model, that provides abstractions to encapsulate these devices, also called resources, enabling developers to interact with them seamlessly. Regarding the \textit{variety of context information and services} issue, we propose a solution for context interpretation that allows developers to create and manage context rules at runtime, and users to set their preferences in the IE. Finally, regarding the issue on the \textit{resource availability}, the framework includes an application interface for the prototyping and management of pervasive applications (IPGAP), focused on visualization and testing UbiApps, mixing real and virtual components.
%in which the basic component has a uniform structure defined as a Resource Agent (RA), formerly proposed in~\cite{cardoso2006} as an entity gathering context information. In this work, the RA, while maintaining context information, acts as a component that encapsulates the device code associated with it, including aspects of interaction with application components. 
%The \textit{heterogeneity of devices} is handled through the definition of a Distributed Component Model, in which the basic component has a uniform structure defined as a Resource Agent (RA), formerly proposed in~\cite{cardoso2006} as an entity gathering context information. In this work, the RA, while maintaining context information, acts as a component that encapsulates the device code associated with it, including aspects of interaction with application components.

The concepts of the framework were materialized on a platform called \textit{SmartAndroid}\footnote{www.tempo.uff.br/smartandroid}, whose development has enabled appraisal in order to prove that the conceptual framework facilitates the process of building UbiApps. We implemented some use cases to validate different aspects of the framework proposed and to testify its capabilities, supported by the elaboration of competency questions, which is a mechanism mostly used to evaluate ontologies, but can also be used as an assessment to this work. 
%power of representation 
%testify the feasibility 
% has enabled the construction of use cases that demonstrate the effectiveness of the approach to provide context awareness to new applications.

% The concepts of the framework have been implemented on a platform called \textit{SmartAndroid}\footnote{www.tempo.uff.br/smartandroid} developed in the context of the project. The implementation of this project has enabled appraisal in order to prove that the conceptual framework facilitates the process of building UbiApps.  One of the evaluations occurred during the processing of an application running on a purely local UbiApp. In another assessment, an application was built to verify the feasibility of implementation of Context Model, through the exploitation of communication mechanisms used in the Distributed Component Model.

The remainder of this paper is organized as follows. First we present in Section~\ref{sec:overview} an overview of the main concepts used as a basis for the framework's development. In Section~\ref{sec:framework}, we present the framework's overall architecture and main features. We then, in Section~\ref{sec:evaluation}, present a proof of concept demonstrating the feasibility of building UbiApps using the framework and we show examples of applications that explores the key features of the IE. Finally, in Section~\ref{sec:related} we compare our approach with related work and in Section~\ref{sec:conclusion} we conclude this paper with the main remarks.
[refactor this paragraph]

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%              SECTION
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work} \label{sec:related}
% [ubicomp is nice but it is not priceless]
The benefits to users and developers that UbiComp foresees reach beyond the horizon of many imaginative researchers. The possibilities that arise from it surpass the bounds established by keyboards and mice, our standard interfaces. Nonetheless, building such adaptable applications requires much effort from developers that see themselves immersed in a universe of device's specifications and communication technologies, apart from the problem itself that the application must solve. Therefore, many researchers have focused on diminishing those obstacles by providing abstractions, middlewares, services, tools, and other supportive techniques, not only for development but also for prototyping of UbiApps.

%ubicomp implies dealing with limited and dynamically varying computational resources [kindberg2002ieee]

Most approaches focus on one aspect of UbiComp, such as security~\cite{}, location~\cite{}, context representation~\cite{}, SmartHome specific~\cite{dixon2012operating}, among others. [+++]
% The works that focus on context interpretation/context rules
% Some works focus on composing and managing context rules  %About the adaptability 
%Challenges in Access Right Assignment for Secure Home Networks - 17

%TODO: Gaia
As one of the most famous, the Gaia middleware is designed to facilitate the construction of applications for IE. It consists of a set of core services and a framework for building distributed context aware applications. The Gaia embraces many goals that include the acquisition of context, the maintenance of hardware and software descriptions, mechanisms to find resources and to compose context rules, among others.
%uses operating systems abstractions to define its architecture
%have many goals that include the provision of mechanisms of acquisition of context; composition of rules (using first logic order), 
%%maintenance of hardwares and softwares description and about applications; 
%search mechanism to find resources
%presence and execution status of resources
%interpretation mechanism that gathers and interprets informations, as defined by programmers
%mvc
%The main goals of Gaia's approach are the acquisition of context by applications, the monitoring of entities location, the maintenance of hardwares and softwares descriptions, and 
%Active Spaces
%FIXME: second phrase was copied

%TODO: Context Toolkit - Dey
%also a famous approach
%we use these abstractions, the aggregation, interpretation, the widget
The Context Toolkit provides a set of abstractions for the composition of reusable components that focus on context acquisition and interpretation. [+++]

%TODO: Gator Tech
[Gator Tech]
%generic reference architecture applicable to any IE 
%middleware layers:
%physical, sensor platform (OSGi , service, knowledge, context management, application

@@

% [explain that our solution strives to accommodate several problems of UbiComp and to be flexible enough to accommodate the others]

The solution presented in this paper strives to accommodate central issues in the development of UbiApps, as identified by the challenges (i), (ii) and (iii). The framework proposed facilitates the construction of those applications but don't restrict its development, being flexible enough to accommodate other concerns that are system specific.
% We propose a framework that

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%              SECTION
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview} \label{sec:overview}
% [context/contx sensibility]
It is easy to notice that the context information is a first concern topic in ubiquitous systems. But what is ``context''? Many authors have proposed definitions to this concept, though most are little accurate~\cite{baldauf2007survey}. Synthesizing previous definitions, Dey and Abowd have proposed that context is ``any information that can be used to characterize the situation of entities (i.e., whether a person, place or object) that are considered relevant to the interaction between a user and an application, including the user and the application themselves''~\cite{dey2001conceptual}. In a IE ``Places'' are the rooms, the floors of a building, or spaces in general that enable the localization of other entities, ``People'' are individuals that populate the environment and interact with it, and ``things'' are virtual representations of physical objects or software components.
%Considering the diversity of devices, systems and protocols, one 
%Many authors have proposed definitions to this concept, though most use examples to explain it or adopt a restrictive definition, that may not contemplate a set of systems. <- wrong. Most are wide or use synonyms
%Synthetizing previous definitions, Dey and Abowd have proposed that context is any relevant information used to characterize the situation of entities, specifically: people, places and things.

Context aware (or context sensible) applications is a designation to those applications that not only are capable of knowing the context of the environment, but also react to it either by means of changing the environment or in a software level. []

[distributed systems]

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%              SECTION
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Framework Architecture} \label{sec:framework}
% [framework description] 
The framework provides facilities and programming standards typically required in context-aware applications focused on IE. Among the features is the ability to abstract details that involve communication, context acquisition and resource location, by the applications. For the sake of abstracting resource details we have used components called \textit{Resource Agents} (RA). The RA is the basic modularization unit of the framework being used in the modeling, implementation and management of applications (Subsection~\ref{subsec:ra}). %Example?

The decentralization of components of the framework (the RAs) follow the principles of the Service-oriented Architecture (SoA), where the more decentralization the more flexibility on connectivity, since the dependence of centralized services is lowered.
%TODO: write it better

% Resolution of the issue is achieved through the decentralization of services, using service-oriented architecture (SoA). The use of SOA enables the relationship between different entities, where the greater decentralization connectivity is more flexible, since the dependence of a service center is minimized. Therefore, problems in a system component are softened and can be absorbed by mechanisms of fault tolerance and persistence.

% Thus, to expose services following the style SoA, other components can access it uniformly reducing the complexity of integration. For example, details of data collection from a temperature sensor are known internally by AR, which is responsible for providing a simple interface to other system components, such as collecting the temperature in Celsius (C) or Fahrenheit (F) that internally, the sensor represented is captured in Kelvin (K).

% [distributed components model]

% The general architecture of the framework is expressed in a form of a distributed components model, which defines the RA structure 
The general architecture of the framework includes a Distributed Components Model, which defines the RA structure and the Resource Management Support Services (RMSS). The Figure~\ref{fig:arch} presents a layer architecture that highlights main aspects of the framework. The layers are described in the following items.

% [architecture picture]
\begin{figure}[!t]
\centering
\includegraphics[width=3.5in]{figures/architecture-refactored-v2.pdf}
\caption{Framework Layer Architecture}
\label{fig:arch}
\end{figure}

\subsubsection{Physical Resources Layer}
The deepest layer is the one where is found the resources present in the environment. In a IE these resources are the smarter versions of daily used objects, such as beds, stoves, TVs, clocks, and so on. Since all these resources have their own mechanisms of communication and operation, then a higher level entity is required to deal with them.

\subsubsection{Middleware Layer}
This layer comprehends the main services that sustain the interoperability of the RAs and provides basic features to the development of UbiApps. The management services and the RA Repository, that compose the middleware, will be further discussed in Subsection~\ref{subsec:sgar}.
% Below the management services a RA Repository keeps , that  as will be further discussed in Subsection~\ref{subsec:sgar},

\subsubsection{Resource Agents Layer}
This layer is composed by all the RAs that represent physical resources already coupled with the IE. The RA will be presented in Subsection~\ref{subsec:ra}.

\subsubsection{Applications Layer}
The higher layer includes all UbiApps that are enjoying the middleware features and also all the software level RAs, which are represented in the figure by the ``Forgotten Stove Interpreter'' and the ``Reminder Actuator''.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%              SECTION
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Framework Features and Application}This section approaches the main features provided by the framework, which includes the definition of the RA (Subsection~\ref{subsec:ra}), the management services (Subsection~\ref{subsec:sgar}) and the context interpretation (Subsection~\ref{subsec:ci}). In the next section we will present the current status of the implementation.
% This section approaches the main features provided by the framework, which includes the definition of the RA (Subsection~\ref{subsec:ra}), the management services (Subsection~\ref{subsec:sgar}) and the context interpretation (Subsection~\ref{subsec:ci}). Also, based on the reference implementation, an application was built to provide a tool for the prototyping, management and visualization of applications and resources running on the IE.

  \subsection{Resource Agents} \label{subsec:ra}
  % [resource]
  % A resource may be defined as any hardware or software entity that exposes services which can be used by other entities and applications present in IE. Thus, sensors, actuators and smart devices (e.g. stove, fridge, air-conditioning) fall within this definition. For example, a light bulb can expose a service to get its current state (either on or off) and another to effectively turn it on and off. A software module that monitors all these light bulbs turned on, deduce the power consumption and expose this information is also an example of resource.

  A resource may be defined as any hardware or software entity that exposes services (of data acquisition or actuation) which can be used by other entities and applications present in IE. Thus, sensors, actuators and smart devices (e.g. stove, fridge, air-conditioning), as well as software modules that provide a service to the environment, fall within this definition.

  The RAs are entities that represent resources. They encapsulate the specificities of resources and expose their information to the environment through standard interfaces, so that others can access them uniformly reducing the complexity of integration.

  The Dey and Abowd's definition to context proposes three entities as the most basic of a ubiquitous system, i.e. person, place and object (or thing). Although the difference of concepts, we have conceived all these three entities as RAs on the framework, but observed their particular features. Thus they inherit the convenience of being a RA.

  % The Resource Agents (RA) can be understood as entities that represent resources, which are sensors, actuators, devices, and smart appliances, as well as software modules that provide a service to the environment. The RAs encapsulate the specificities of resources and expose their information to the environment through standard interfaces, so that others can access them uniformly reducing the complexity of integration. For example, data collection details from a temperature sensor would be known only by their RA, which is responsible for providing a simple interface to other system components.
  %it's similar to the widget of the context toolkit
  %enjoys benefits of SOA

  %From gator tech:
  %decoupling sensors and actuators from sensor platforms ensures openness and makes it possible to introduce new technology as it becomes available

  [wrapper]

    \subsubsection{RA Interfaces}

    The RA architecture includes two interfaces for input and output. The former is called Context Variable (CV) and is responsible for exposing the RA's context information. The latter is called Operation (OP) and is responsible for exposing features (or actuation services) belonging to RAs, enabling applications to actively interact in the environment.
    % In the previous example, the ``oven is turned on'' and the ``oven temperature in ºC'' are the some of the RA's CVs.

    Consider the simple example of a smart stove where the instantiation of its RA exposes services to get its current context (e.g., oven is turned on, oven temperature in $\,^{\circ}\mathrm{C}$) and services that may change its context (e.g., turn oven off, set temperature as $150\,^{\circ}\mathrm{C}$). Hence, the RA concept meets the first challenge (i), which is the \textit{devices heterogeneity}, since decoupling sensor and actuator from the sensor platforms favors the flexibility. %FIXME: is it right?

    % The RA are distributed components that can be used to

    \subsubsection{Communication}

    % [communication mechanisms]
    % The communication performed between two different ARs or between an AR and an application is developed through '
    % Towards implementing 
    In order to meet requirements typical of distributed environments, two communications mechanisms have been included in the framework: the Remote Procedure Call (RPC), that implements synchronous communication; and a publish-subscribe mechanism (pub-sub)~\cite{eugster2003many}, that implements asynchronous communication. The RPC is used to access directly the RA, querying for its context by accessing its CVs, or calling its OPs. On the other hand, through the pub-sub a RA can subscribe to a CV of another one (consumer) and further receive a notification, when the first RA publishes its CV state change (provider).

    The RA has a domain name, the RA Name System (RANS), that allows its unique identification in the IE. This identification is similar to the DNS, where the name remains the same even if the network address (IP) changes, the only change is the updating of the addressing table.

    \subsubsection{RA Structure}
    In analogy to object-oriented programming we define classes of RAs to represent the type of the resource with specific interfaces. Besides other informations present in the AR's structure (such as name, location, subscribed RAs, among others) it has its type hierarchy. The type hierarchy is composed by a sequence of classes that classify the entity and has been inspired by the ontology described by Ranganathan~\cite{ranganathan2005olympus}. In the Subsection~\ref{subsec:sgar} we will present how this feature can be used to find a RA and to define its functionalities.

  \subsection{Resource Management Support Service} \label{subsec:sgar}
  A set of services is responsible for providing basic functionalities to manage the RAs. The four main services are the RA Register Service (RRS), the RA Discovery Service (RDS), the RA Location service (RLS) and the Security Management Service (SecMS). All these services access the RA Repository (RR) that maintains informations about the RAs (addresses and descriptions) and the map data structure (map of the environment). The Figure~\ref{fig:arch} represents, in the Middleware layer, the RR and above it the management services.

  The management services are also designed as RAs, hence they take advantage of the communication mechanisms already provided by it and communicate using the same primitives.

    \subsubsection{RA Register Service} 
    The RRS allows the registration and deregistration of RAs instances references and description. This service also has CVs that allows the developer to know when RAs enter and leave the IE

    The naming and addressing scheme can be generalized to work over the internet, and comply with the Internet of Things concept.

    \subsubsection{RA Discovery Service}
    The RDS allows the search for registered RAs based on specific attributes, which are the RA name, RANS and type. The search by RANS will only return one RA, but the search by name or type will return all RAs that meet this criteria.

    After the discovery of a RA, the stakeholder can instantiate a stub (i.e. a proxy) of this RA, when both the stub and the current implementation of the RA implement the same interface. Through the stub the CVs and OPs can be accessed, what internally will be converted to network calls to the RA object (RPC functioning).

    During the application initialization the only information required get started is the RDS address. And even this information can be obtained, if within a LAN, by sending a broadcast message with the request; or using the DNS, if within a WAN. With the RDS address on hands any RA can be found, including the other management services (i.e., the RRS, RLS and RSS). 
    %, before doing the connections to RAs,

    \subsubsection{RA Location Service}
    The RLS explores the map of the environment to run the required queries. This service performs searches related to the location of RAs, such as the physical location of RAs, the RAs located on some place, the RAs of some type located on some place, the RAs close to another and order by proximity, among others. Thus, if a developer needs audiovisual devices close to some person, for example, she can run this search through the RLS and will receive a list, where an utility function can choose the more relevant result.

    The RLS exposes CVs that allows not only the query for if a RA is located at some place, but also to subscribe to that place for the sake of being notified of the entrance or exit of a specified RA type. Therefore, even if the RA is still not registered, an application can subscribe to a place and wait for its appearance.

    Nowadays there are several technologies to map and/or track the location of entities, that may include from people to small things. These technologies vary mainly based on precision, ubiquity (capable of being ``invisible'') and price. As examples we have the GPS, the smart floor, cameras (associated to image processing softwares), Wi-Fi triangulation, RFID tags, and so on. In this approach we do not comply with any specific technology, and we assume that the RA can discover its own location or the RLS can.
    % already established technology of location

    \subsubsection{Security Management Service}
    The security concern is always a central topic on UbiComp systems, moreover because of diverse aspects of security required. Some aspects are already inherited depending on the technology adopted, e.t., the Wi-Fi already ensures against the registration of outside devices and the exchange of informations.

    [credentials]

    [domains]
    % The SecMS manages both the user's credentials and the application's domains.
    % 
    % On the

  \subsection{Context Interpretation} \label{subsec:ci}
  [motivation]

  % [directly programmed]
  The context interpretation serves to aggregate context information from different sources in accordance with some specific logic and considering the passage of time. This logic is defined by the developer and the interpretation is implemented with the framework's support. %The CI concept was inspired by the Aggregator in Dey and Abowd~\cite{dey2001conceptual} whose objective was to gather logically related context and provides them in a single component.

  The entity that performs the context interpretation may be encapsulated by a RA, thus allowing the subscription to other RAs, what is called Context Interpreter (CI). This concept allows dealing with context in a higher level and also provides a separation on concerns, since a context interpreter decouples the steps for context acquisition, evaluation, timing and notification, that are repeatedly performed by the UbiApps. Therefore, using the CIs also promotes the reuse, either by the creator application or by other ones. 
  %FIXME: Doubt: ``a RA'' or ``an RA''?

  The framework uses a generic approach for creating context rules, which are composed by two parts: the interpretation and the actuation. The interpretation part, as already stated before, concerns on binding the CI itself with CVs from different RAs, evaluating the expression that interrelates those CVs, and notifying whoever is interested. The actuation part is a RA that is subscribed to the CI, and develop some job, that may change the context of the IE or change some state in a software level.
%   in which developers are able to define the logic of the rule in the application using the framework support for context acquisition, as well as to perform actions on the environment.

  An example of context rules has been presented on Figure~\ref{fig:arch}. At the Applications Layer a set of two components implements the interpretation-actuation pattern: the ``Forgotten Stove Interpreter'' and the ``Reminder Actuator'' repectivelly, where the goal of the rule is to remind the elderly that he has forgotten the oven on while he went to sleep. The former component, which is a CI, is subscribed to two RAs, namely the ``Bed RA'' and the ``Stove RA'' (the dashed arrow means notify subscriptors). The latter is subscribed to the CI and calls the RAs, namely the ``TV RA'' and the ``Clock RA'', directly using RPC (represented by the straight arrow). Internally to the CI a rule expression relates the RAs as  ``bed is occupied and oven is turned on for 30 min'', which means that the CI only will be valid if this condition remains true during ``30 min''. When the CI changes its state from invalid to valid (what includes the timing), the subscriptors are notified, the ``Reminder Actuator'' in case. This actuator, in its turn, calls the OPs ``fire alarm'' from ``Clock RA'' and ``set message'' from ``TV RA'', with the parameter ``You forgot the oven on!''. Then it is expected that an accident is prevented.

%   In this simple example, the CI is subscribed to the RAs ``Bed RA'' and ``Stove RA'', and internally it contains the following rule expression: ``bed is occupied and stove is on for T min''. The actuator, awaits for when the condition of the CI met for triggering its action that, in case, is to play the clock's alarm sound and to show on TV a message saying that ``the stove has been forgotten''.
%   The Figure~\ref{fig:arch} shows on Applications Layer an example of a context rule with two components: the CI (``Forgotten Stove Interpreter'') and the actuator (``Reminder Actuator'').

%   A avaliação da expressão é desempenhada pelo próprio IC quando o contexto muda. A atuação ocorre quando um agente atuador é subscrito à um IC, recebendo assim a notificação de que a expressão da regra foi avaliada como verdadeira. Estes agentes atuadores podem tanto ser atuadores simples, que desempenham ações sequencialmente no ambiente utilizando OPs diretamente (e.g., desligar lâmpada, gravar programação de determinado canal da TV); como atuadores mais complexos, que executam ações não necessariamente baseadas em OPs (e.g., execução de algum outro software, que utiliza um canal de emergência do serviço de saúde ou bombeiros).

% [talk more about actuators]
  An actuator can be simple as a list of actions to be invoked sequentially, such as in the previous example, or it can be more complex, e.g. a program that send a SMS message to some person in case of emergency, or that consults the whether from an website and shows on a visual interface at morning, and many others. %In Subsection~\ref{subsec:rules} we will present an easier way for programming simple actuators, and also how to implement more complex actuators.
%   program that records TV program if person is late from work

%   [can have many actuators]
  The independence of the actuator allows its binding and unbinding with some CI in a easily manner, what is make possible because of the RA communication mechanism. Based on the same principle, more than one actuator can be bound to the same CI, for example, we may bind the forgotten stove CI with a ``family warning actuator'', that contacts a relative by phone and put her touch with the elderly.

%   [timer can be CV]
  Another aspect of context rules is the time dimension. The definition of timers inside a rule expression can be done by subscribing the CI to a Timer RA, hence enabling the direct reference in the expression. The timer can also be relative or absolute. An absolute timer is objective: if the timer has reached some timestamp, or if it is within a window of time, then it is valid. [+++]

  % [ COMO É A TEMPORIZAÇÃO NO IC ]
% A temporização em um IC é definida ao se associar um ou mais temporizadores a uma subexpressão da regra (ou a expressão completa). Os temporizadores podem ser tanto relativos (e.g., fogão ligado e pessoa deitada na cama por 15min) quanto absolutos (e.g., ar-condicionado ligado às 18:00).

% [ INTERRUPÇÃO DE TEMPORIZADOR E AVALIAÇÃO DE REGRAS TEMPORIZADAS ]
% Durante a avaliação, se a expressão ou uma subexpressão que possuir um temporizador associado chegar em um estado verdadeiro, o temporizador é inicializado. Se, durante uma avaliação, a expressão ou subexpressão se torna falsa, então os temporizadores associados são interrompidos. Se o tempo limite é alcançado e se não há outros temporizadores associados a subexpressões da mesma regra em andamento, então os atuadores são notificados.
